---
title: "Measles susceptibility in Glasgow"
number-sections: true
format: pdf
editor_options: 
  chunk_output_type: console
theme: flatly
---

```{r}
#| echo: false
#| warning: false
#| message: false
library(ggplot2)
library(tidyverse)
library(sjPlot)
library(performance)
library(kableExtra)
library(knitr)
```

## Introduction

Measles is a highly contagious viral disease that can lead to serious health complications, especially in young children. Vaccination programmes such as the measles, mumps, and rubella (MMR) vaccine have been highly effective in reducing the spread of measles. However, in public confidence in vaccination programmes can strongly influence vaccination uptake and therefore the level of susceptability within a population.

In 1998, a study by Wakefield et al. suggested a potential link between the MMR vaccine and autism. Although this claim was later widely discredited and the paper was partially retracted in 2004, the resulting media attention led to a notable decline in vaccination uptake across parts of the United Kingdom. Reduced vaccination coverage increases the proportion of children susceptible to measles, potentially increasing risk of outbreaks.

This report analyses measles susceptibility among pre-school children in Glasgow between 1998 and 2014 using data from the Scottish Childhood Immunisation Record System (SCIRS). The main objective is to investigate whether measles susceptibility changed over time and whether any change occurred following the retraction of the Wakefield article in 2004. In particular, the analysis focuses on modelling the proportion of children susceptible to measles across the study period.

## Exploratory Analysis

```{r}
#| echo: false
#| message: false

glasmeasles <- read.csv("DAProject13.csv") |>
  mutate(prop = Y/N)
```

To understand the trajectory of measles susceptibility in Glasgow between 1998 and 2014, we first examined the proportion of susceptible pre-school children (Y/N) rather than just raw counts. This allows for fair comparison across the 133 intermediate zones.

To observe the broader city-wide trend, we aggregated these data to calculate the overall annual proportion of susceptible children across Glasgow. @fig-trend visualizes this progression over the 16-year period, with a dashed line at 2004 to help visually assess if the article's partial retraction aligned with a clear change in susceptibility levels.

```{r}
#| echo: false
#| fig-cap: Measles Susceptibility in Glasgow Pre-schoolers (1998-2014). The red dashed line indicates the 2004 partial retraction of the Wakefield article
#| label: fig-trend
#| fig-align: center
#| fig-width: 4
#| fig-height: 2

yearly_trend <- glasmeasles %>%
  group_by(Year) %>%
  summarise(
    Total_Y = sum(Y),
    Total_N = sum(N),
    Proportion = Total_Y / Total_N
  )


ggplot(yearly_trend, aes(x = Year, y = Proportion)) +
  geom_line(color = "skyblue", linewidth = 1) +
  geom_point(color = "darkblue", size = 2) +
  geom_vline(xintercept = 2004, linetype = "dashed", color = "red", linewidth = 1) +
  labs(
    x = "Year",
    y = "Proportion of Susceptible Children"
  ) +
  theme_minimal() +
  scale_x_continuous(breaks = seq(1998, 2014, by = 2))
```

@fig-trend demonstrates a clear shift in measles susceptibility. Between 1998 and 2004, the proportion of susceptible children rose to a peak of just over 15%. A distinct reversal in this trend occurred alongside the 2004 article retraction. Following this 2004 turning point, susceptibility levels steadily declined, reaching their lowest observed rate of approximately 8% by the end of the study period.

To understand if the city-wide trend reflects all of Glasgow or just a few areas, we must look at the 133 individual neighbourhoods. @fig-box illustrates the yearly spread of susceptibility across these zones. This helps us see how much the neighbourhoods differ from one another and identifies any outliers with abnormally high susceptibility.

```{r}
#| echo: false
#| warning: false
#| fig-cap: "Distribution of measles susceptibility rates across Glasgow's 133 intermediate zones (1998-2014)."
#| label: fig-box
#| fig-align: center
#| fig-pos: "H"
#| fig-width: 4
#| fig-height: 2

glasmeasles %>%
  ggplot(aes(x = factor(Year), y = prop)) +
  geom_boxplot(fill = "steelblue4", alpha = 0.6) +
  labs(
    x = "Year",
    y = "Proportion of Susceptible Children"
  ) +
  theme_minimal()
```

As seen in @fig-box, Glasgow's neighborhood medians peak in 2004, following the city-wide trend. Variability was highest between 2002 and 2006, suggesting the vaccine scare affected some areas much more than others. By 2014, neighbourhoods became more similar as susceptibility fell city-wide. Persistent outliers highlight specific high-risk zones, such as one neighborhood in 2000 with a rate exceeding 45%, well above the city average.

To support the visual trends observed in the previous figures, @tbl-summary provides the precise numerical values for the mean, median, and standard deviation of measles susceptibility across the 133 intermediate zones for each year.

```{r}
#| echo: false
#| label: tbl-summary
#| tbl-cap: Mean, median and standard deviation (sd) of the proportion of measles susceptibility in Glasgow by year.
#| tbl-pos: "H"

summary <- glasmeasles %>%
  summarise('Prop_Mean' = mean(prop, na.rm = TRUE),
            'Prop_Median' = median(prop, na.rm = TRUE),
            'Prop_sd' = sd(prop, na.rm = TRUE),
            .by = Year)

summary %>%
  kable(col.names = c("Year",
                      "Mean Proportion",
                      "Median Proportion",
                      "Proportion SD"),
        digits = 3,
                table.attr = "style='width: 15%; font-size: 5px; '")
```

The mean and median proportions in @tbl-summary confirm that susceptibility peaked in 2004 at roughly 15%. Furthermore, the Standard Deviation (SD) column measures the variability discussed earlier. The SD was highest between 2000 and 2002, confirming that the gap between neighbourhoods was widest during the early years of the vaccine scare. By 2012, the SD fell to its lowest level ($0.024$), showing that susceptibility rates across Glasgow became more uniform as they declined.

## Formal Analysis

```{r}
#| echo: false
#| message: false

glasmeasles <- read.csv("DAProject13.csv") |>
  mutate(prop = Y/N, post_2004 = as_factor(ifelse(Year >= 2004, 1, 0)))
glasmeasles$Year <- as.factor(glasmeasles$Year)
```

Building on the patterns observed in the exploratory analysis, logistic regression models were fitted to formally assess changes in measles susceptibility over time, using year as a categorical predictor and as a binary indicator for pre and post 2004.

```{r}
#| echo: false
#| message: false
#| label: tbl-yearmodel
#| tbl-cap: Coefficients, CIs and p-values for the logistic model treating year as a categorial predictor.

year_model <- glm(prop ~ Year, data = glasmeasles, weights = N, family = binomial)

year_model %>% tab_model(transform = NULL)
```

```{r}
#| echo: false
#| message: false

check_overdispersion(year_model)
```

```{r}
#| echo: false
#| message: false
#| fig-cap: Predicted probability of a Glasgow pre-schooler being susceptible to measles by year (1998-2014). The vertical line around points indicates the 95% confidence interval.
#| label: fig-yearpred

plot_model(type = "pred",
           year_model,
           terms = "Year",
           axis.title = c("Year",
                          "Prob. of a child being susceptible to measles"),
           title = " ")
```

The year-specific model confirms the temporal pattern identified in the exploratory analysis. Relative to 1998, susceptibility increases significantly between 2000 and 2004, with the largest effects observed in 2002 and 2004 (log-odds $\approx$ 0.33-0.34, p \< 0.001), corresponding to probabilities of around 15-16%. Although 2006 remains elevated, the effect is smaller, indicating the start of a decline. From 2008 onwards, coefficients become negative, with significant reductions by 2010 and continuing through 2012 and 2014. This closely aligns with the earlier visual trends, providing strong statistical evidence for a peak in the early 2000s followed by a sustained decline.
The overdispersion test is statistically significant (dispersion ratio = 1.156, p \< 0.001), but the magnitude is small, suggesting the model remains appropriate. 

To formally assess whether a structural change occurred around 2004, a second model compares observations before and after this point.

```{r}
#| echo: false
#| message: false
#| label: tbl-binarymodel
#| tbl-cap: Coefficients, CIs and p-values for the logistic model treating year as a binary predictor (observations from before 2004/observations from 2004 and later).

prepost_model <- glm(prop ~ post_2004, data = glasmeasles, weights = N, family = binomial)

prepost_model %>% tab_model(transform = NULL)
```

```{r}
#| echo: false
#| message: false
#| fig-cap: Predicted probability of a Glasgow pre-schooler being susceptible to measles for before 2004 (0) or 2004 and later (1). The vertical line around points indicates the 95% confidence interval.
#| label: fig-binarypred

plot_model(type = "pred",
           prepost_model,
           terms = "post_2004",
           axis.title = c("Before and after 2004",
                          "Prob. of a child being susceptible to measles"),
           title = " ")
```

The pre/post 2004 model tests the apparent turning point identified in the exploratory analysis. The post 2004 coefficient is negative and highly significant (log-odds = -0.23, p \< 0.001), indicating a reduction in susceptibility after 2004. This corresponds to a decrease in probability from approximately 13.5% to 11%, consistent to a decrease with the observed downward trend. This provides strong evidence of a structural change in susceptibility occurring around 2004.

Overall, the results provide consistent evidence that susceptibility increased in the early 2000s, peaked around 2002-2004, and declined significantly thereafter. The agreement between both models strengthens the conclusion that a meaningful change occurred around 2004.

## Conclusions

Overall, the analysis provides clear evidence that measles susceptibility in Glasgow did change following the retraction of the Wakefield article. Susceptibility increased from 1998 to the early 2000s, reaching a peak of around 15% in 2002–2004, before declining steadily to approximately 8% by 2014.

The formal modelling results provide strong statistical evidence that this change is significant. Our year-specific logistic regression model shows significantly higher log-odds of susceptibility in the early 2000s compared to the baseline year, followed by significantly lower values in later years. More importantly, the pre/post 2004 model provides strong evidence of a structural change, with a negative coefficient for the post-2004 indicator, indicating a decrease in susceptibility around 2004 and confirming it as a turning point. This is consistent with the timing of the Wakefield article retraction.

However, the study is limited by the use of aggregated data and the absence of additional explanatory variables, meaning that other factors such as public health interventions or socioeconomic differences may also explain the observed trend. As a result, while the timing of the change coincides with the Wakefield article retraction, the analysis cannot establish a causal relationship. In future studies, incorporating additional explanatory variables or extending the model to account for spatial and temporal dependence would better explain the drivers of measles susceptibility.
